\documentclass[12pt,fleqn]{report}
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{amsmath}
\usepackage{natbib}
\usepackage{longtable}
\usepackage{appendix}
\usepackage{mathtools}
\usepackage{hyperref}
\usepackage{tikz}

% Workaround for longtable labels
\AtBeginDocument{%
\newlabel{tab:rentdata}{{}{}}%
\newlabel{tab:prepostdata}{{}{}}%
\newlabel{tab:bushdata}{{}{}}%
\newlabel{tab:outdata}{{}{}}%
}


\title{Notes on regression analysis for social research}
\author{E. C. Hedberg \\ hedbergec@gmail.com}
\date{}
\newcommand\intro[1]{
 \chapter*{#1}
 \markboth{\MakeUppercase{#1}}{}
 \addcontentsline{toc}{chapter}{#1}
}

\begin{document}
\maketitle
\tableofcontents
\listoftables
\listoffigures
\intro{About}
This document is a collection of my notes for an advanced regression course. It comes with no assurance of absolute accuracy. These notes are updated constantly and dated.

NOTE: These notes are based on the work of many statisticians and social scientists, and in no way reflect my own intellectual contributions.

I will eventually build a larger bibliography, but for right now these are just notes for a class. These notes are not designed to replace a proper text. ~\nocite{*}

I assume you have had a basic regression course and, of course, introduction to statistics. This is not about theory, and I go through the basics quickly.

\input{chUniv.tex}

\input{chProb.tex}

\input{chSRS.tex}

\input{chDist.tex}

\input{chInfer.tex}

\input{chMulti.tex}

\input{chLSML.tex}

\input{chOLS.tex}

\input{chMOLS.tex}

\input{chOLSinfer.tex}

\input{chColin.tex}

\input{chOut.tex}

\input{chCoding.tex}

\input{chANCOVA.tex}

\input{chInter.tex}

\input{chHetero.tex}

\input{chGLS.tex}

\input{chLogit.tex}

\bibliographystyle{plainnat}
\bibliography{bib}

\end{document} 