
\chapter{Univariate statistics}

\section{The mean, statistics, and parameters}
To calculate the mean, simply add up the numbers for each of the cases and divide by the total number of cases. For example, the mean of the numbers 1,2,3,4,5,6,7,8,9 is (1+2+3+4+5+6+7+8+9) / 9 = 41/9 = 4.556. Statisticians and mathematicians symbolize the addition of values in a set by using the summation symbol, $\Sigma$. The $\sum_{i=1}^{N}$ phrase represents the summation of each $i^{th}$ case, starting with the first one $i = 1$ and ending with the $N^{th}$ case. To calculate the mean, this sum is divided by $N$, or the total number of cases.

The formula of the mean is:
\begin{equation}\label{eq:mean}
\mu=\frac{\sum_{i=1}^{N}x_i}{N},
\end{equation}
which is estimated by the statistic:
\begin{equation}
\bar{x}=\frac{\sum_{i=1}^{N}x_i}{N}.
\end{equation}
In the previous paragraph, the term statistic is used. It is important to note that there are two concepts: parameters and statistics, which are often conflated by students. Frequentists generally think that the population at large can be described with parameters. For example, one parameter is the mean, or $\mu$. For example, a statistician might say that the population of objects (e.g., people) has some mean income. The statistician could then use a sample of objects to estimate this parameter. This estimate is the statistic.

For example, the estimate of the population mean, $\mu$, of some variable {\it x} is the mean statistic $\bar{x}$. If the variable was {\it w}, the statistic would be $\bar{w}$. The estimate of the population variance $\sigma^2$ is the statistic $s^2$. One way to think about this is with the mnemonic {\bf PPSS}: {\underline P}opulations have {\underline P}arameters, {\underline S}amples have {\underline S}tatistics.

Of course, these statistics are not going to be absolutely correct, and will vary from one sample to the next. Thus, one of the things that researchers fret most about is how well a certain statistic estimates a parameter. This is the sampling variation of that statistic, or the variance of the estimate. The square root of the variance of a statistic is the standard error. More on that later.
\section{Deviance}
The next important idea is that of deviance. This is central to most of what statistics is about since statisticians often wonder how one population differs from another that has experienced some effect or treatment. The most basic deviation is the difference between a case's value of {\it x} and the mean of {\it x}, $\bar{x}$. Deviance can be symbolized as:
\begin{equation}
  e_i=x_i-\bar{x}.
\end{equation}
Even this basic idea has substantive interest. It is often important to know how some case differs from what is typical. Here, the mean is considered typical, and it is valuable to know whether a case is typical, above typical, or below typical. Negative deviance values are below the mean, positive deviance values are above the mean, and 0 deviance is exactly the mean.
\section{The sum of squares}
It is often important to understand how much deviance there is in our sample. One method might be to add up each deviance score, but because of properties of the mean this will wind up at 0 which makes this a rather useless method. As luck would have it, there is a simple solution to this problem.

By squaring each deviance, the negative values of deviance are effectively made positive (since the square of any negative number is positive) and then by adding up each squared deviance term, a measurement of total deviance is obtained. We call this total deviance the sum of squares, or $SS$:
\begin{equation}
  SS=\sum_{i=1}^{N}e_i^2={\sum_{i=1}^{N}(x_i-\bar{x})^2}.
\end{equation}
One interesting way to view $SS$ is to consider it a measure of how much information is available in a variable. The larger the $SS$, the more information that is contained in the variable that can be modeled and correlated with other variables. If the $SS$ is small, then little will be correlated with it.
\section{The variance}
The next quantity is the variance. This is simply the average of $SS$. In the population, this is
\begin{equation}
  \sigma_x^2=\frac{{\sum_{i=1}^{N}(x_i-\bar{x})^2}}{N}.
\end{equation}
The estimate of the population variance is the sum of squared deviations divided by N-1:
\begin{equation}
  s_x^2=\frac{{\sum_{i=1}^{N}(x_i-\bar{x})^2}}{N-1}.
\end{equation}
We divide by N-1 instead of N because we lose a degree of freedom since this statistic employs another statistic (the mean). This is also a correction to remove bias, since this estimate of the population variance without this correction is slightly smaller than the expected parameter.
\section{The standard deviation}
A problem with the variance is that it in units of squared deviations. To get back to simple deviations, we take the square root of the variance to get the standard deviation. In the population this is
\begin{equation}
  \sigma_x=\sqrt{\frac{{\sum_{i=1}^{N}(x_i-\bar{x})^2}}{N}},
\end{equation}
and the statistic is
\begin{equation}
  s_x=\sqrt{\frac{{\sum_{i=1}^{N}(x_i-\bar{x})^2}}{N-1}}.
\end{equation}

\section*{For more information}
For foundational understanding of univariate statistics, see \citep{rice2007mathematical}, \citep{agresti}, and \citep{skk}.